---
title: "Econometrics. Pair-wise linear regression"
subtitle: "Discussion 3-2: The Crimea Reunion"
author: "Evgenii V. Gilenko (e.gilenko@gsom.spbu.ru)"
output: html_document
---

# 1. Initial manipulations

R Markdown language allows to fully introduce the power of **literate programming**.

In this chunk the necessary libraries are activated. No warning messages reflected.

```{r loadlib, message=F, warning=F}
library(openxlsx) # to import data from MS Excel files
library(Rlab) #  hplot
library(car) # qqPlot; linear hypotheses testing
library(DescTools) # RMSE, MAE, MAPE, SMAPE
```

Let's check and, if necessary, set the working directory to correctly import the data from a file.

Now, let's import the data to a data-frame. If the file is in the **same directory**, then we just need to specify the name of the file.

```{r}
keynes <- read.xlsx("RussianGDP.xlsx", sheet="data")
```


# 2. Preliminary data analysis

First, let's calculate some summary statistics for the dataset. You can apply the *summary* command right to the data-frame.

```{r}
summary(keynes)
```

Now, let's construct individual graphs for the two time-series.

```{r}
## Individual graph for CONS
plot(x=keynes$year, y=keynes$cons, type="b", lwd=3, col="blue", pch=19 , main="Households' consumption expenditures", xlab="Years", ylab="Consumption expenditures (bln rub.)"); grid()
```


```{r}
## Individual graph for GDP
plot(x=keynes$year, y=keynes$gdp, type="b", lwd=3, col="blue", pch=19 , main="Gross Domestic Product", xlab="Years", ylab="GDP (bln rub.)"); grid()
```


Since we study the interaction of the two time-series, let's construct a scatter plot for them. The scatter plot actually involves several interesting tricks.


```{r}
## set up the colors
keynes$color <- 1
keynes$color[13] <- 2

# construct the plot
plot(x=keynes$gdp, y=keynes$cons, pch=19, col=keynes$color, ylim=c(8000,28000), main="Consumption vs GDP", xlab="GDP (bln rub.)", ylab="Consumption (bln rub.)", cex=1); grid()

# add textual labels
text(x=keynes$gdp, y=keynes$cons+1000, labels=keynes$year, cex=1, pos=3, col=keynes$color) 
```

Let's not forget about correlation analysis.

```{r}
cor.test(keynes$cons, keynes$gdp)
```


# 3. Pair-wise regression: Model 1.

Let's estimate our Model 1 (**without the final observation!**).

```{r}
## estimate the model
model1 <- lm(data=keynes[keynes$year!=2014, ], cons~gdp)
summary(model1)
```

```{r}
## change the format of numbers
options(scipen = 4)
```



## 3.1. Hypotheses testing for Model 1.

Here we will run some formal tests of Model 1.

### 3.1.1. Testing normality of the disturbances

```{r}
# histogram
hplot(model1$residuals, main = "Residuals of Model 1")
```

```{r}
# Q-Q plot
qqPlot(model1$residuals, distribution="norm", id=F, pch=19,
       main="Q-Q plot for the residuals of Model 1")
```

```{r}
# the Shapiro-Wilk test
shapiro.test(model1$residuals)
```


### 3.1.2. Linear restrictions

```{r}
# a function from package car
linearHypothesis(model1, c("gdp=1")) 
```


## 3.2. Forecasting with Model 1

### 3.2.1 Goodness-of-fit

Let's analyze goodness-of-fit of the model.

```{r}
# information criteria
AIC(model1)
BIC(model1)
```

```{r}
# traditional prediction metrics
RMSE(model1)
MAE(model1)
```

```{r}
# percentage prediction metrics
MAPE(model1)
SMAPE(model1)
```

### 3.2.2 Making predictions

First, we construct a one-step ahead forecast with a **confidence** interval (at the 95% confidence level).

```{r}
predict.lm(model1, newdata=keynes[keynes$year==2014, ], interval="confidence", level=0.95)
```

Then, we construct a one-step ahead forecast with a **predictive** interval (at the 95% confidence level).

```{r}
predict.lm(model1, newdata=keynes[keynes$year==2014, ], interval="prediction", level=0.95)
```


# 4. Pair-wise regression: Model 2

Let's transform the model.

```{r}
# subtract the mean of GDP
keynes$dmgdp <- keynes$gdp - mean(keynes$gdp[keynes$year!=2014])
```

```{r}
# estimate the new version of the model
model2 <- lm(data=keynes[keynes$year!=2014, ], cons~dmgdp)
summary(model2)
```





